Nilpotent Invariants \\ for Generic Discrete Series of Real Groups

### 2025-12-22 revision_whittaker.tex
Adds discussion of the action on Whittaker data

\subsubsection*{Whittaker models}

Suppose again that $\pi$ is an irreducible admissible representation of $G(\R)$ on a Hilbert space $\mH$. We consider the corresponding actions of $G(\R)$ on the space~$\mH^\infty$ of smooth vectors in $\mH$, and on its dual $\mH^{-\infty}$. Let~$(B, \eta)$ be a pair satisfying conditions (i)-(ii) of the Introduction. We say $\pi$ has a $(B,\eta)$-Whittaker model
if there exists a nonzero element $v$ in~$\mH^{-\infty}$ such that $\pi(n)(v)=\eta(n)v$ for all $n$ in the nilradical $N(\R)$ of $B(\R)$. See~\cite{matumoto}. \textcolor{red}{[NEW]} This condition depends only on the $G(\R)$-conjugacy class of the pair $(B, \eta)$. It is equivalent to the existence of a $G$-equivariant embedding of $\mH^\infty$ into the space of smooth sections of the equivariant line bundle over $G(\R)/N(\R)$ determined by $\eta$ (for the quasi-regular action of~$G$ on sections); the image of that embedding is a geometric realization of (the smooth vectors of)~$\pi$ which is the origin of the terminology ``Whittaker model for~$\pi$''. \textcolor{red}{[/NEW]}

 If~$\w$ is the corresponding Whittaker datum, i.e. the $G(\R)$-conjugacy class of the pair $(B, \eta)$, we say~$\pi$ has a $\w$-Whittaker model.

@@

Recall that the \emph{Weyl chambers} of~$\mathfrak{t}^\ast$ are the connected components of the set of regular elements. Each Weyl chamber specifies a positive root system for the roots of $\mathfrak{t}$ in~$\g$, \textcolor{red}{[new]} and a Borel subgroup of~$G$ containing~$T$ and defined over~$\R$ \textcolor{red}{[/new]}. We say a Weyl chamber is \emph{large} if the corresponding simple roots are all noncompact.

@@

\textcolor{red}{[NEW]} For the action on~(1) see Section~\ref{sec:action_DS}. For the action on~(2) and the equivariance of $\pi \mapsto \Wh(\pi)$, see Section~\ref{sec:action_whitt}. For the action on~(4) and the equivariance of the map $\pi \mapsto \AV(\pi)$, see Section~\ref{sec:action_AV}. For the action on~(3) and the equivariance of $\pi \mapsto \WF(\pi)$ see Section~\ref{sec:action_WF}. \textcolor{red}{[/NEW]}

@@

\begin{remarkplain} \textcolor{red}{(NEW)} \label{rk:size_Q} When $G$ is adjoint, it immediately follows from the definitions that $(G_\ad)^\sigma = (G^\sigma)_{\ad}$, therefore $Q_\sigma(G)$ is trivial in that case.

@@

\begin{center} \textcolor{red}{TO DO : explain that this preserves generic representations, and  DS representations within a single $L$-packet.\\ Then explain why the action is transitive.   }\end{center}

\subsubsection{\textcolor{red}{(NEW)} On Whittaker data}\label{sec:action_whitt}

\begin{center} \textcolor{red}{TO DO : explain how $Q_\sigma(G)$ acts on Whittaker data, and refer to \cite[Lemma 3.1.3]{adams_kaletha} for the simple transitivity; point out that $\pi \mapsto \Wh(\pi)$ is equivariant.   }\end{center}

>>>>

\subsubsection*{Whittaker models}\label{sec:whittaker_defs}

Suppose again that $\pi$ is an irreducible admissible representation of $G(\R)$ on a Hilbert space $\mH$. We consider the corresponding actions of $G(\R)$ on the space~$\mH^\infty$ of smooth vectors in $\mH$, and on its dual $\mH^{-\infty}$. Let~$(B, \eta)$ be a pair satisfying conditions (i)-(ii) of the Introduction; we shall say $(B, \eta)$ is a \emph{Whittaker pair}. We say $\pi$ has a \emph{$(B,\eta)$-Whittaker model}
if there exists a nonzero element $v$ in~$\mH^{-\infty}$ such that $\pi(n)(v)=\eta(n)v$ for all $n$ in the nilradical $N(\R)$ of $B(\R)$. See~\cite{matumoto}. \textcolor{red}{[NEW]} This condition depends only on the $G(\R)$-conjugacy class of the pair $(B, \eta)$. It is equivalent to the existence of a $G$-equivariant embedding of $\mH^\infty$ into the space of smooth sections of the equivariant line bundle over $G(\R)/N(\R)$ determined by $\eta$ (for the quasi-regular action of~$G$ on sections); the image of that embedding is a geometric realization of (the smooth vectors of)~$\pi$ which is the origin of the terminology ``Whittaker model for~$\pi$''. \textcolor{red}{[/NEW]}

 If~$\w$ is the corresponding Whittaker datum, i.e. the $G(\R)$-conjugacy class of the pair $(B, \eta)$, we say~$\pi$ has a \emph{$\w$-Whittaker model}.

@@

Recall that the \emph{Weyl chambers} of~$\mathfrak{t}^\ast$ are the connected components of the set of regular elements. Each Weyl chamber specifies a positive root system for the roots of $\mathfrak{t}$ in~$\g$, \textcolor{red}{[new]} and a Borel subgroup of~$G$ containing~$T$ and defined over~$\R$ \textcolor{red}{[/new]}. We say a Weyl chamber is \emph{large}, or that the corresponding Borel subgroup is large, if the corresponding simple roots are all noncompact.

@@

\textcolor{red}{[NEW]} For the action on~(1) see Sections~\ref{sec:action_DS} and~\ref{sec:action_transitivity}. For the action on~(2) and the equivariance of $\pi \mapsto \Wh(\pi)$, see Section~\ref{sec:action_whitt}. For the action on~(4) and the equivariance of the map $\pi \mapsto \AV(\pi)$, see Section~\ref{sec:action_AV}. For the action on~(3) and the equivariance of $\pi \mapsto \WF(\pi)$ see Section~\ref{sec:action_WF}. \textcolor{red}{[/NEW]}

@@

\begin{remarkplain} \textcolor{red}{[NEW]} \label{rk:size_Q} When $G$ is adjoint, it immediately follows from the definitions that $(G_\ad)^\sigma = (G^\sigma)_{\ad}$, therefore $Q_\sigma(G)$ is trivial in that case.

@@

\subsubsection{\textcolor{red}{(NEW SEC.)} On Whittaker data and generic representations}\label{sec:action_whitt}

Suppose $\tilde{q} \in \Gad^\sigma$. For every Whittaker pair $(B, \eta)$ as in Section~\ref{sec:whittaker_defs}, the pair $q\cdot (B, \eta)=\big(\tilde{q} B \tilde{q}^{-1}, \eta \circ (x \mapsto \tilde{q} x \tilde{q}^{-1})\big)$ is again a Whittaker pair, which is conjugate to $(B, \eta)$ if and only if $\tilde{q} \in (G^\sigma)_{\ad}$. This determines an action of $Q_\sigma(G)$ on the set of Whittaker data for~$G(\R)$, which we shall denote by $(q, \mathfrak{w}) \mapsto q \cdot \mathfrak{w}$. This action is simply transitive: see \cite[Lemma 3.1.3]{adams_kaletha} for an elementary argument.

Now let $\pi$ be a generic discrete series representation of~$G(\R)$. Let $q$ be an element of $Q_\sigma(G)$, with a representative $\tilde{q}$ in $\Gad^\sigma$.  For every Whittaker pair $(B, \eta)$ such that $\pi$ has a $(B, \eta)$-Whittaker model, it is immediate from the definitions that the representation $\pi \circ (x \mapsto \tilde{q} x\tilde{q}^{-1})$ is again a discrete series, and that it has a $\tilde{q}\cdot (B, \eta)$-Whittaker model. Passing to equivalence classes we see that $\pi^q$ is generic, and satisfies $\Wh(\pi^q)=q \cdot \Wh(\pi)$. Therefore the map $\pi \mapsto \Wh(\pi)$ is $Q_\sigma(G)$-equivariant.

\subsubsection{On generic discrete series in a single $L$-packet}\label{sec:action_transitivity}

\begin{center} \textcolor{red}{TO DO : explain that the action of $Q_\sigma$ on generic DS preserves $L$-packets.\\ Then explain why the action on is transitive.\\
Perhaps adding something about $L$-packets in \S~\ref{sec:discrete_series} would be useful. }\end{center}
